---
output:
  rmarkdown::github_document
bibliography: "inst/REFERENCES.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# R/`medoutcon`

[![Travis-CI Build Status](https://travis-ci.org/nhejazi/medoutcon.svg?branch=master)](https://travis-ci.org/nhejazi/medoutcon)
[![AppVeyor Build  Status](https://ci.appveyor.com/api/projects/status/github/nhejazi/medoutcon?branch=master&svg=true)](https://ci.appveyor.com/project/nhejazi/medoutcon)
[![Coverage Status](https://img.shields.io/codecov/c/github/nhejazi/medoutcon/master.svg)](https://codecov.io/github/nhejazi/medoutcon?branch=master)
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](http://opensource.org/licenses/MIT)

> Efficient Causal Mediation Analysis with Mediator-Outcome Confounding

__Authors:__ [Nima Hejazi](https://nimahejazi.org), [Iván
Díaz](https://idiaz.xyz), and [Kara
Rudolph](http://biostat.jhsph.edu/~krudolph/)

---

## What's `medoutcon`?

The `medoutcon` R package provides facilities for efficient estimation of
stochastic (in)direct effects that measure the impact of a treatment variable
$A$ on an outcome variable $Y$, through a direct path (through $A$ only) and
an indirect path (through a set of mediators $M$ only), in the presence of an
intermediate <b>med</b>iator-<b>out</b>come <b>con</b>founder $Z$, itself
affected by the treatment $A$. While the proposed approach is similar to those
appearing in @vanderweele2014effect, @rudolph2017robust, and
@zheng2017longitudinal, `medoutcon` is designed as a software implementation to
accompany the methodology proposed in @diaz2019nonparametric. Both an efficient
one-step bias-corrected estimator with cross-fitting
[@pfanzagl1985contributions; @zheng2011cross; @chernozhukov2018double] and a
one-step cross-validated targeted minimum loss (TML) estimator
[@vdl2011targeted; @zheng2011cross] are made available. `medoutcon` integrates
with the [`sl3` R package](https://github.com/tlverse/sl3) [@coyle2019sl3] to
leverage statistical machine learning in the estimation procedure.
<!--
and expands on
the architecture exposed by the [`tmle3` R
package](https://github.com/tlverse/tmle3) for TML estimation.
-->
---

## Installation

Install the most recent _stable release_ from GitHub via
[`remotes`](https://CRAN.R-project.org/package=remotes):

```{r gh-master-installation, eval=FALSE}
remotes::install_github("nhejazi/medoutcon")
```

---

## Example

To illustrate how `medoutcon` may be used to estimate stochastic (in)direct
effects of the treatment (`A`) on the outcome (`Y`) in the presence of
mediator(s) (`M`) and a binary mediator-outcome confounder (`Z`), consider the
following simple example:

```{r simple_example, message=FALSE, warning=FALSE}
library(stringr)
library(data.table)
library(medoutcon)

# produces a simple data set based on ca causal model with mediation
make_example_data <- function(n_obs = 1000) {
  # baseline covariate -- simple, binary
  W <- replicate(2, rbinom(n_obs, 1, prob = 0.50))
  W <- as.data.table(W)
  setnames(W, c("w_1", "w_2"))

  # create treatment based on baseline W
  A <- as.numeric(rbinom(n_obs, 1, prob = rowSums(W)/3 + 0.1))

  # single mediator-outcome confounder
  z_prob <- 1 - plogis((A^2 + rowMeans(W)) / (A + rowSums(W^3) + 0.5))
  Z <- rbinom(n_obs, 1, prob = z_prob)

  # matrix of mediators
  m1_prob <- plogis((A^2 - Z) / (A + rowMeans(W) + 0.5))
  m2_prob <- 1 - plogis((A + rowSums(W)) / (Z + 0.5))
  M1 <- rbinom(n_obs, 1, prob = m1_prob)
  M2 <- rbinom(n_obs, 1, prob = m2_prob)
  M <- as.data.table(list(m_1 = M1, m_2 = M2))

  # create outcome as a linear function + white noise
  Y <- rowMeans(M) - Z + A - 0.1 * rowSums(W) +
    rnorm(n_obs, mean = 0, sd = 0.25)

  # full data structure
  data <- as.data.table(cbind(Y, M, Z, A, W))
  return(data)
}

# set seed and simulate example data
set.seed(75681)
example_data <- make_example_data()
w_names <- str_subset(colnames(example_data), "w")
m_names <- str_subset(colnames(example_data), "m")

# compute one-step estimate
os_medoutcon <- medoutcon(W = example_data[, ..w_names],
                          A = example_data$A,
                          Z = example_data$Z,
                          M = example_data[, ..m_names],
                          Y = example_data$Y,
                          estimator = "onestep",
                          estimator_args = list(cv_folds = 3))
summary(os_medoutcon)

# compute targeted minimum loss estimate
tmle_medoutcon <- medoutcon(W = example_data[, ..w_names],
                            A = example_data$A,
                            Z = example_data$Z,
                            M = example_data[, ..m_names],
                            Y = example_data$Y,
                            estimator = "tmle",
                            estimator_args = list(cv_folds = 3))
summary(tmle_medoutcon)
```

For details on how to use data adaptive regression (machine learning) techniques
in the estimation of nuisance parameters, consider consulting the vignette that
accompanies the package.

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/nhejazi/medoutcon/issues).

---

## Contributions

Contributions are very welcome. Interested contributors should consult our
[contribution
guidelines](https://github.com/nhejazi/medoutcon/blob/master/CONTRIBUTING.md)
prior to submitting a pull request.

---

## Citation

After using the `medoutcon` R package, please cite the following:

        @article{diaz2019nonparametric,
          title={Non-parametric efficient causal mediation with intermediate
            confounders},
          author={D{\'\i}az, Iv{\'a}n and Hejazi, Nima S and Rudolph, Kara E
            and {van der Laan}, Mark J},
          year={2019+},
          url = {},
          doi = {},
          journal={},
          volume={},
          number={},
          pages={},
          publisher={}
        }

        @manual{hejazi2019medoutcon,
          author={Hejazi, Nima S and D{\'\i}az, Iv{\'a}n and Rudolph, Kara E},
          title = {{medoutcon}: Efficient causal mediation analysis with
            intermediate confounding},
          year  = {2019},
          url = {https://github.com/nhejazi/medoutcon},
          note = {R package version 0.0.4}
        }

---

## License

&copy; 2019 [Nima S. Hejazi](https://nimahejazi.org)

The contents of this repository are distributed under the MIT license. See below
for details:
```
MIT License

Copyright (c) 2019 Nima S. Hejazi

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in all
copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
SOFTWARE.
```

---

## References

