---
title: "Efficient causal mediation analysis with intermediate confounders"
author: "[Nima Hejazi](https://nimahejazi.org), [Iván
  Díaz](https://www.idiaz.xyz/), and [Kara
  Rudolph](https://kararudolph.github.io/)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Efficient causal mediation analysis with intermediate confounders}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Background and Motivations

An exposure of interest often affects an outcome directly, or indirectly by the
mediation of some intermediate variables. Identifying and quantifying the
mechanisms underlying causal effects is an increasingly popular endeavor in
public health, medicine, and the social sciences, as knowledge of such
mechanisms can improve understanding of both _why and how_ treatments can be
effective. Such mechanistic knowledge may be arguably even more important in
cases where treatments result in unanticipated ineffective or even harmful
effects.

Traditional techniques for mediation analysis fare poorly in the face of
intermediate confounding. Classical parameters like the natural (in)direct
effects face a lack of identifiability in cases where mediator-outcome (i.e.,
intermediate) confounders affected by exposure complicate the relationship
between the exposure, mediators, and outcome. @diaz2020nonparametric provide a
theoretical and computational study of the properties of newly developed
interventional (in)direct effect estimands within the non-parametric statistical
model. Among their contributions, @diaz2020nonparametric

- derive the efficient influence function (EIF), an key object in
  semiparametric efficiency theory;
- use the EIF to develop two asymptotically optimal, non-parametric estimators,
  each of which is capable of leveraging machine learning for the estimation of
  nuisance parameters; and
- present theoretical conditions under which their proposed estimators are
  consistent, multiply robust, and efficient.

# Problem Setup and Notation

The problem addressed by the work of @diaz2020nonparametric may be represented
by the following nonparametric structural equation model (NPSEM):
\begin{align*}
  W &= f_W(U_W); A = f_A(W, U_A); Z=f_Z(W, A, U_Z);\\ \nonumber
  M &= f_M(W, A, Z, U_M); Y = f_Y(W, A, Z, M, U_Y).
\end{align*}
In the NPSEM, $W$ denotes a vector of observed pre-treatment covariates, $A$
denotes a categorical treatment variable, $Z$ denotes an intermediate confounder
affected by treatment, $M$ denotes a (possibly multivariate) mediator, and $Y$
denotes a continuous or binary outcome.  The vector of exogenous factors
$U=(U_W,U_A,U_Z,U_M,U_Y)$, and the functions $f$, are assumed deterministic but
unknown.  Importantly, the NPSEM encodes a time-ordering between these variables
and allows the evaluation of counterfactual quantities defined by intervening on
a set of nodes of the NPSEM.  The observed data unit can be represented by the
random variable $O = (W, A, Z, M, Y)$; we consider access to $O_1, \ldots, O_n$,
a sample of $n$ i.i.d. observations of $O$.

@diaz2020nonparametric additionally define the following parameterizations,
familiarity with which will be useful for using the [`medoutcon` `R`
package](https://github.com/nhejazi/medoutcon). In particular, these authors
define $g(a \mid w)$ as the probability mass function of $A = a$ conditional on
$W = w$ and use $h(a \mid m, w)$ to denote the probability mass function of $A
= a$ conditional on $(M, W) = (m, w)$. Further, @diaz2020nonparametric use
$b(a, z, m, w)$ to denote the outcome regression function $\mathbb{E}(Y \mid A
= a, Z = z, M = m, W = w)$, as well as $q(z \mid a,w)$ and $r(z \mid a, m, w)$
to denote the corresponding conditional densities of $Z$.

# Interventional (In)Direct Effects

@diaz2020nonparametric define the _total effect_ of $A$ on $Y$ in terms of a
contrast between two user-supplied values $a', a^{\star} \in \mathcal{A}$.
Examination of the NPSEM reveals that there are four paths involved in this
effect, namely $A \rightarrow Y$, $A \rightarrow M \rightarrow Y$, $A
\rightarrow Z \rightarrow Y$, and $A \rightarrow Z \rightarrow M \rightarrow Y$.
Mediation analysis has classically considered the _natural direct effect_ (NDE)
and the _natural indirect effect_ (NIE), which are defined as
$\mathbb{E}_c(Y_{a', M_{a^{\star}}} - Y_{a^{\star}, M_{a^{\star}}})$ and
$\mathbb{E}_c(Y_{a',M_{a'}} - Y_{a',M_{a^{\star}}})$, respectively. The natural
direct effect measures the effect through paths _not_ involving the mediator
($A \rightarrow Y$ and $A \rightarrow Z \rightarrow Y$), whereas the natural
indirect effect measures the effect through paths involving the mediator
($A \rightarrow M \rightarrow Y$ and $A \rightarrow Z \rightarrow M \rightarrow
Y$). As the sum of the natural direct and indirect effects equals the average
treatment effect $\mathbb{E}_c(Y_1-Y_0)$, this effect decomposition is
appealing. Unfortunately, the natural direct and indirect effects are
not generally identified in the presence of an intermediate confounder affected
by treatment.

To circumvent this issue, @diaz2020nonparametric define the direct and indirect
effects using stochastic interventions on the mediator, following a strategy
previously outlined by @vanderweele2014effect and @rudolph2017robust, among
others. Let $G_a$ denote a random draw from the conditional distribution of
$M_a$ conditional on $W$. Consider the effect of $A$ on $Y$ defined as the
difference in expected outcome in hypothetical worlds in which $(A,M) = (a',
G_{a'})$ versus $(A,M) = (a^{\star}, G_{a^{\star}})$ with probability one, which
may be decomposed into direct and indirect effects as follows
\begin{equation*}
\mathbb{E}_c(Y_{a', G_{a'}} - Y_{a^{\star}, G_{a^{\star}}}) =
  \underbrace{\mathbb{E}_c(Y_{a', G_{a'}} - Y_{a',
    G_{a^{\star}}})}_{\text{Indirect effect (through $M$)}} +
  \underbrace{\mathbb{E}_c(Y_{a', G_{a^{\star}}} - Y_{a^{\star},
      G_{a^{\star}}})}_{\text{Direct effect (not through $M$)}}.
\end{equation*}
Like the natural direct effect, this interventional direct effect measures the
effects through paths not involving the mediator. Likewise, the interventional
indirect effect measures the effect through paths involving the mediator. Note,
however, that natural and interventional mediation effects have different
interpretations. That is, the interventional indirect effect measures the effect
of fixing the exposure at $a'$ while setting the mediator to a random draw
$G_{a^{\star}}$ from those with exposure $a'$ versus a random draw $G_{a'}$ from
those with exposure $a^{\star}$, given covariates $W$. As is clear from the
effect decomposition, the term $\theta_c = \mathbb{E}_c(Y_{a', G_{a^{\star}}})$
is required for estimation of both the interventional direct and indirect
effects; thus, @diaz2020nonparametric focus on estimation of this quantity.
Importantly, it has been shown that $\theta_c$ is identified by the statistical
functional
\begin{equation*}
  \theta = \int b(a', z, m, w) q(z \mid a', w) p(m \mid a^{\star}, w)
    p(w) d\nu(w,z,m)
\end{equation*}
under a set of standard identifiability conditions [@vanderweele2014effect],
which are further reviewed in @diaz2020nonparametric.

# Efficient Estimation

@diaz2020nonparametric define two efficient estimators of their interventional
(in)direct effects. These are based on the one-step estimation and targeted
minimum loss (TML) estimation frameworks, respectively. Briefly, both estimation
strategies proceed in two stages, starting by first constructing initial
estimates of the nuisance parameters present in the EIF, then proceeding to
apply distinct bias-correction strategies in their second stages. Both
estimation strategies require an assumption about the behavior of initial
estimators of the nuisance parameters (specifically, that these lie in a Donsker
class); however, the need for such an assumption may be avoided by making use of
cross-validation in the fitting fo initial estimators. The `medoutcon` `R`
package requires the use of cross-validation in the construction of these
initial estimates, resulting in cross-fitted one-step and and cross-validated
TML estimators [@klaassen1987consistent; @zheng2011cross;
@chernozhukov2018double].

The one-step estimator $\hat{\theta}_{\text{os}}$ is constructed by adding the
empirical mean of the EIF (evaluated at initial estimates of the nuisance
parameters) to the substitution estimator. By constrast, the TML estimator
$\hat{\theta}_{\text{tmle}}$ updates the components of the substitution
estimator via logistic tilting models formulated to ensure that relevant score
equations appearing in the EIF are (approximately) solved.  While the estimators
are asymptotically equivalent, TML estimators have been shown to exhibit
superior finite-sample performance, making them potentially more reliable than
one-step estimators. For the exact form of the EIF as well as those of the
one-step and TML estimators, consult @diaz2020nonparametric.

# Data Analysis Example

## Setting up the data example

Now, we'll take a look at how to estimate the interventional direct and indirect
effects using a simulated data example. @diaz2020nonparametric illustrate the
use of their estimators of these effects in an application in which they seek to
elucidate the mechanisms behind the unintended harmful effects that a housing
intervention had on adolescent girls' risk behavior.

First, let's load a few required packages and set a seed for our simulation.

```{r setup, message=FALSE, warning=FALSE}
library(data.table)
library(medoutcon)
library(sl3)
set.seed(75681)
n_obs <- 500
```

Next, we'll generate a very simple simulated dataset. The function
`make_example_data`, defined below, generates three binary baseline covariates
$W = (W_1, W_2, W_3)$, a binary exposure variable $A$, a single binary mediateor
$M$ an intermediate confounder $Z$ that affects the mediator $M$ and is itself
affected by the exposure $A$, and, finally, a binary outcome $Y$ that is a
function of $(W, A, Z, M)$.

```{r make_example_data, message=FALSE, warning=FALSE}
# produces a simple data set based on ca causal model with mediation
make_example_data <- function(n_obs = 1000) {
  ## baseline covariates
  w_1 <- rbinom(n_obs, 1, prob = 0.6)
  w_2 <- rbinom(n_obs, 1, prob = 0.3)
  w_3 <- rbinom(n_obs, 1, prob = pmin(0.2 + (w_1 + w_2) / 3, 1))
  w <- cbind(w_1, w_2, w_3)
  w_names <- paste("W", seq_len(ncol(w)), sep = "_")

  ## exposure
  a <- as.numeric(rbinom(n_obs, 1, plogis(rowSums(w) - 2)))

  ## mediator-outcome confounder affected by treatment
  z <- rbinom(n_obs, 1, plogis(rowMeans(-log(2) + w - a) + 0.2))

  ## mediator -- could be multivariate
  m <- rbinom(n_obs, 1, plogis(rowSums(log(3) * w[, -3] + a - z)))
  m_names <- "M"

  ## outcome
  y <- rbinom(n_obs, 1, plogis(1 / (rowSums(w) - z + a + m)))

  ## construct output
  dat <- as.data.table(cbind(w = w, a = a, z = z, m = m, y = y))
  setnames(dat, c(w_names, "A", "Z", m_names, "Y"))
  return(dat)
}

# set seed and simulate example data
example_data <- make_example_data(n_obs)
w_names <- stringr::str_subset(colnames(example_data), "W")
m_names <- stringr::str_subset(colnames(example_data), "M")
```

Now, let's take a quick look at our simulated data:

```{r example_data, message=FALSE, warning=FALSE}
# quick look at the data
head(example_data)
```

As noted above, all covariates in our dataset are binary; however, note that
this need not be the case for using our methodology --- in particular, the only
current limitation is that the intermediate confounder $Z$ must be binary when
using our implemented TML estimator of the (in)direct effects.

Using this dataset, we'll proceed to estimate the interventional (in)direct
effects. In order to do so, we'll need to estimate several nuisance parameters,
including the exposure mechanism $g(A \mid W)$, a re-parameterized exposure
mechanism that conditions on the mediators $h(A \mid M, W)$, the outcome
mechanism $b(Y \mid M, Z, A, W)$, and two variants of the intermediate
confounding mechanism $q(Z \mid A, W)$ and $r(Z \mid M, A, W)$. In order to
estimate each of these nuisance parameters flexibly, we'll rely on data adaptive
regression strategies in order to avoid the potential for (parametric) model
misspecification.

<!--
Note that there are two additional nuisance parameters that must also be
estimated ($u$ and $v$), which are themselves functions of the other nuisance
parameters.  We recommend estimating these via the highly adaptive lasso, which
is the...
-->

## Ensemble learning of nuisance functions

As we'd like to rely on flexible, data adaptive regression strategies for
estimating each of the nuisance parameters $(g, h, b, q, r)$, we require a
method for choosing among or combining the wide variety of available regression
strategies. For this, we recommend the use of the Super Learner algorithm for
ensemble machine learning [@vdl2007super].  The recently developed [`sl3` R
package](https://tlverse.org/sl3) [@coyle2020sl3] provides a unified interface
for deploying a wide variety of machine learning algorithms (simply called
_learners_ in the `sl3` nomenclature) as well as for constructing Super Learner
ensemble models of such learners. For a complete guide on using the `sl3` R
package, consider consulting https://tlverse.org/sl3, or https://tlverse.org
(and https://github.com/tlverse) for the `tlverse` ecosystem, of which `sl3` is
an integral part.

To construct an ensemble learner using a handful of popular machine learning
algorithms, we'll first instantiate variants of learners from the appropriate
classes for each algorithm, and then create a Super Learner ensemble via the
`Lrnr_sl` class. Below, we demonstrate the construction of an ensemble learner
based on a modeling library including an intercept model, a main-terms GLM,
$\ell_1$-penalized Lasso regression, an elastic net regression that equally
weights the $\ell_1$ and $\ell_2$ penalties, random forests (`ranger`), and the
highly adaptive lasso (HAL):

```{r make_sl, message=FALSE, warning=FALSE}
# instantiate learners
mean_lrnr <- Lrnr_mean$new()
fglm_lrnr <- Lrnr_glm_fast$new(family = binomial())
lasso_lrnr <- Lrnr_glmnet$new(alpha = 1, family = "binomial", nfolds = 3)
enet_lrnr <- Lrnr_glmnet$new(alpha = 0.5, family = "binomial", nfolds = 3)
rf_lrnr <- Lrnr_ranger$new(num.trees = 200)

# for HAL, use linear probability formulation, with bounding in unit interval
hal_gaussian_lrnr <- Lrnr_hal9001$new(
  family = "gaussian",
  fit_control = list(
    max_degree = 3,
    n_folds = 3,
    use_min = TRUE,
    type.measure = "mse"
  )
)
bound_lrnr <- Lrnr_bound$new(bound = 1e-6)
hal_bounded_lrnr <- Pipeline$new(hal_gaussian_lrnr, bound_lrnr)

# create learner library and instantiate super learner ensemble
lrnr_lib <- Stack$new(mean_lrnr, fglm_lrnr, enet_lrnr, lasso_lrnr,
                      rf_lrnr, hal_bounded_lrnr)
sl_lrnr <- Lrnr_sl$new(learners = lrnr_lib, metalearner = Lrnr_nnls$new())
```

While we recommend the use of a Super Learner ensemble model like the one
constructed above in practice, such a library will be too computationally
intensive for our examples. To reduce computation time, we construct a simpler
library, using only a subset of the above learning algorithms:

```{r make_simple_sl, message=FALSE, warning=FALSE}
# create simpler learner library and instantiate super learner ensemble
lrnr_lib <- Stack$new(mean_lrnr, fglm_lrnr, lasso_lrnr, rf_lrnr)
sl_lrnr <- Lrnr_sl$new(learners = lrnr_lib, metalearner = Lrnr_nnls$new())
```

Having set up our ensemble learner, we're now ready to estimate each of the
interventional effects using the efficient estimators exposed in the `medoutcon`
package.

## Estimating the direct effect

We're now ready to estimate the interventional direct effect. This direct effect
is computed as a contrast between the interventions $(a' = 1, a^{\star} = 0)$
and $(a' = 0, a^{\star} = 0)$. In particular, our efficient estimators of the
interventional direct effect proceed by constructing estimators
$\hat{\theta}(a' = 1, a^{\star} = 0)$ and $\hat{\theta}(a' = 0, a^{\star} = 0)$.
Then, an efficient estimator of the direct effect is available by application
of the delta method, that is, $\hat{\theta}^{\text{DE}} =
\hat{\theta}(a' = 1, a^{\star} = 0) - \hat{\theta}(a' = 0, a^{\star} = 0)$.
Applying the same principle to the EIF estimates, one can derive variance
estimates and construct asymptotically correct Wald-style confidence intervals
for $\hat{\theta}^{\text{DE}}$.

The `medoutcon` package makes the estimation task quite simple, as only a single
call to the eponymous `medoutcon` function is required. As demonstrated below,
we need only feed in each component of the observed data $O = (W, A, Z, M, Y)$
(of which $W$ and $M$ can be multivariate), specify the effect type, and the
estimator. Additionally, for each nuisance parameter we may specify a separate
regression function --- in the examples below, we use the simpler Super Learner
ensemble constructed above for fitting each nuisance function, but this need not
be the case (i.e., different estimators may be used for each nuisance function).

First, we examine the one-step estimator of the interventional direct effect.
Recall that the one-step estimator is constructed by adding the mean of the EIF
(evaluated at initial estimates of the nuisance parameters) to the substitution
estimator. As noted above, this is done separately for each of the two contrasts
$(a' = 0, a^{\star} = 0)$ and $(a' = 1, a^{\star} = 0)$. Thus, the one-step
estimator of this direct effect is constructed by application of the delta
method to each of the one-step estimators (and EIFs) for these contrasts.

```{r de_os, message=FALSE, warning=FALSE}
# compute one-step estimate of the interventional direct effect
os_de <- medoutcon(W = example_data[, ..w_names],
                   A = example_data$A,
                   Z = example_data$Z,
                   M = example_data[, ..m_names],
                   Y = example_data$Y,
                   g_learners = sl_lrnr,
                   h_learners = sl_lrnr,
                   b_learners = sl_lrnr,
                   q_learners = sl_lrnr,
                   r_learners = sl_lrnr,
                   effect = "direct",
                   estimator = "onestep",
                   estimator_args = list(cv_folds = 2))
summary(os_de)
```

From the output of the summary method, we note that the one-step estimate of
the interventional direct effect $\hat{\theta}_{\text{os}}^{\text{DE}}$ is
`r round(os_de$theta, 3)`, with 95% confidence interval
[`r round(unname(confint(os_de, 0.95)[1]), 3)`,
`r round(unname(confint(os_de, 0.95)[3]), 3)`].

Next, let's compare the one-step estimate to the TML estimate. Analogous to the
case of the one-step estimator, the TML estimator can be evaluated via a single
call to the `medoutcon` function:

```{r de_tmle, message=FALSE, warning=FALSE}
# compute targeted minimum loss estimate of the interventional direct effect
tmle_de <- medoutcon(W = example_data[, ..w_names],
                     A = example_data$A,
                     Z = example_data$Z,
                     M = example_data[, ..m_names],
                     Y = example_data$Y,
                     g_learners = sl_lrnr,
                     h_learners = sl_lrnr,
                     b_learners = sl_lrnr,
                     q_learners = sl_lrnr,
                     r_learners = sl_lrnr,
                     effect = "direct",
                     estimator = "tmle",
                     estimator_args = list(cv_folds = 2, max_iter = 5))
summary(tmle_de)
```

From the output of the summary method, we note that the TML estimate of the
interventional direct effect $\hat{\theta}_{\text{tmle}}^{\text{DE}}$ is
`r round(tmle_de$theta, 3)`, with 95% confidence interval
[`r round(unname(confint(tmle_de, 0.95)[1]), 3)`,
`r round(unname(confint(tmle_de, 0.95)[3]), 3)`]. Here, we recall that the TML
estimator generally exhibits better finite-sample performance than the one-step
estimator [@vdl2011targeted; @vdl2018targeted], so the TML estimate is likely to
be more reliable in our modest sample size of $n =$ `r n_obs`.

## Estimating the indirect effect

Estimation of the interventional indirect effect proceeds similarly to the
strategy discussed above for the corresponding direct effect. An efficient
estimator can be computed as a contrast between the interventions $(a' = 1,
a^{\star} = 0)$ and $(a' = 1, a^{\star} = 1)$. Specifically, our efficient
estimators of the interventional indirect effect proceed by constructing
estimators $\hat{\theta}(a' = 1, a^{\star} = 0)$ and $\hat{\theta}(a' = 1,
a^{\star} = 1)$.  Then, application of the delta method yields an efficient
estimator of the indirect effect, that is, $\hat{\theta}^{\text{IE}} =
\hat{\theta}(a' = 1, a^{\star} = 0) - \hat{\theta}(a' = 1, a^{\star} = 1)$. The
same principle may be applied to the EIF estimates to derive variance estimates
and construct asymptotically correct Wald-style confidence intervals for
$\hat{\theta}^{\text{IE}}$.

Now, we examine the one-step estimator of the interventional indirect effect.
The one-step estimator is constructed by adding the mean of the EIF
(evaluated at initial estimates of the nuisance parameters) to the substitution
estimator. As noted above, this is done separately for each of the two contrasts
$(a' = 1, a^{\star} = 1)$ and $(a' = 1, a^{\star} = 0)$. Thus, the one-step
estimator of this indirect effect is constructed by application of the delta
method to each of the one-step estimators (and EIFs) for the contrasts.

```{r ie_os, message=FALSE, warning=FALSE}
# compute one-step estimate of the interventional indirect effect
os_ie <- medoutcon(W = example_data[, ..w_names],
                   A = example_data$A,
                   Z = example_data$Z,
                   M = example_data[, ..m_names],
                   Y = example_data$Y,
                   g_learners = sl_lrnr,
                   h_learners = sl_lrnr,
                   b_learners = sl_lrnr,
                   q_learners = sl_lrnr,
                   r_learners = sl_lrnr,
                   effect = "indirect",
                   estimator = "onestep")
summary(os_ie)
```

From the output of the summary method, we note that the one-step estimate of
the interventional indirect effect $\hat{\theta}_{\text{os}}^{\text{IE}}$ is
`r round(os_ie$theta, 3)`, with 95% confidence interval
[`r round(unname(confint(os_ie, 0.95)[1]), 3)`,
`r round(unname(confint(os_ie, 0.95)[3]), 3)`].

As before, let's compare the one-step estimate to the TML estimate. Analogous
to the case of the one-step estimator, the TML estimator can be evaluated via a
single call to the `medoutcon` function, as demonstrated below

```{r ie_tmle, message=FALSE, warning=FALSE}
# compute targeted minimum loss estimate of the interventional indirect effect
tmle_ie <- medoutcon(W = example_data[, ..w_names],
                     A = example_data$A,
                     Z = example_data$Z,
                     M = example_data[, ..m_names],
                     Y = example_data$Y,
                     g_learners = sl_lrnr,
                     h_learners = sl_lrnr,
                     b_learners = sl_lrnr,
                     q_learners = sl_lrnr,
                     r_learners = sl_lrnr,
                     effect = "indirect",
                     estimator = "tmle")
summary(tmle_ie)
```

From the output of the summary method, we note that the TML estimate of the
interventional indirect effect $\hat{\theta}_{\text{tmle}}^{\text{IE}}$ is
`r round(tmle_ie$theta, 3)`, with 95% confidence interval
[`r round(unname(confint(tmle_ie, 0.95)[1]), 3)`,
`r round(unname(confint(tmle_ie, 0.95)[3]), 3)`]. As before, the TML estimator
provides better finite-sample performance than the one-step estimator, so it may
be preferred in this example.

## References

